Towards Practical First-Order Model Counting

### 2024-08-11 doc/paper/paper.tex
The text is all there. Now I just need to shorten it and do some editing.

% TODO: Having an algorithm that can efficiently handle large domain sizes is
% crucial in the weighted setting. For example, consider the well-known "friends
% and smokers" Markov logic network [1]. The utility of such a model would be
% significantly limited if exact probabilities could only be efficiently
% computed for social networks of at most 30 people.
% [1] Richardson, M., & Domingos, P. (2006). Markov logic networks. Machine learning, 62, 107-136.

% TODO: (somewhere) The implementation is only for unweighted (but the
% definition of WFOMC is still necessary because of Skolemization).

\section{Introduction}
% TODO: make it extra clear what the gap is that this work is trying to fill

% TODO: mention Crane earlier and describe its importance

% TODO: the last two paragraphs should make more of a BOOM (e.g., include the
% over 9k claim)

% TODO (here and elsewhere): clarify the distinction between the weighted and
% unweighted variants of the problem in the final version of the paper. We focus
% on the unweighted variant in our implementation and experiments, however,
% Crane2 can handle WFOMC instances as well simply by switching the main data
% type from integers to rational numbers. Maybe add some weighted experiments as
% well.

% TODO: (must) more related work, situating my contributions in the broader
% context

% TODO: (concerning related work) the topic is intimately related to "graded
% logics", largely investigated in DL, mu-calculus and temporal logics

@@

\emph{First-order model counting} is the task of counting the number of models
of a sentence in first-order logic over some given domain(s). The (symmetric)
weighted variant of this problem, known as WFOMC, seeks to calculate the sum of
model weights. In WFOMC, the weight of a model is determined by predicate
weights~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC is a key approach to
\emph{lifted (probabilistic) inference}, which aims to compute probabilities
more efficiently by leveraging symmetries inherent in the problem
description~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By considering domains of
increasing sizes, the model count of a formula can be seen as an integer
sequence. WFOMC algorithms have been utilised for discovering new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} as well as
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. Additionally, WFOMC
algorithms have been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

@@

The complexity of WFOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the WFOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}.
\citeauthor{DBLP:conf/pods/BeameBGS15}~\shortcite{DBLP:conf/pods/BeameBGS15}
demonstrated the existence of an unliftable formula with three variables. It is
also known that all formulas with up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

There are various WFOMC algorithms with different underlying principles. Perhaps
the most prominent class of WFOMC algorithms is based on \emph{first-order
  knowledge compilation} (FOKC). In this approach, the formula is compiled into
a representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
WFOMC for any combination of domain sizes (or weights). Algorithms in this class
include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles
formulas into circuits, while the latter compiles them first to graphs and then
to (algebraic) equations. Another WFOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

% TODO: introduce MLNs somewhere? maybe use an acronym?
The recently proposed \textsc{Crane} algorithm, while capable of handling
formulas beyond the reach of \textsc{ForcLift}, was `incomplete' as it could
only construct function definitions, which would then need to be evaluated to
compute the WFOMC.\@ In other words, it could not be used as a black box that
takes Markov logic networks (or weighted formulas in some other format) and
outputs numbers (i.e., model counts or probabilities). Additionally, recursive
functions were presented without base cases. In this work, we introduce
\Cranetwo{}, an extension of \textsc{Crane} that addresses these two weaknesses.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a
WFOMC problem are compiled into C++ programs. Note that a solution to WFOMC that
uses compilation to C++ has been considered
before~\cite{DBLP:conf/kr/KazemiP16}, however, the extent of formulas that could
be handled was limited. Finally, \cref{sec:experiments} presents experiments
results comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

In \cref{sec:logic}, we provide a summary of the basic principles of first-order
logic. Then, in \cref{sec:threelogics}, we formally define WFOMC and discuss the
distinctions between three variations of first-order logic that are utilised for
WFOMC.\@ Finally, in \cref{sec:algebra}, we introduce the terminology used to
describe the output of the original \textsc{Crane} algorithm, i.e., functions
and equations that define them.

@@

\subsection{WFOMC Algorithms and Their Logics}\label{sec:threelogics}

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in WFOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in WFOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by the most recent implementation
  of
  \textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}.
\end{enumerate*}
The notation we use to refer to each logic is standard in the case of \Ctwo{}
and \UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be
more specific in the case of \FO{}. All three logics are function-free, and
domains are always assumed to be finite. As usual, we presuppose the
\emph{unique name assumption}, which states that two constants are equal if and
only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

@@

@@

It is worth noting that these base cases overlap when $l = m = 0$ but remain
consistent since $0^{0} = 1$. Generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the WFOMC
of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
WFOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{function call $f(\mathbf{y})$ on the right-hand side of an equation in $\mathcal{E}$}{\label{line:functioncall}
    $\mathbf{x} \gets \text{the parameters of $f$ in its definition}$\;
    \ForEach{$y_{i} \in \mathbf{y}$} {
      \uIf{$y_{i} \in \mathbb{N}_{0}$}{
        $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$\;
      }
      \ElseIf{$y_{i} = x_{i} - c_{i}$ for some $c_{i} \in \mathbb{N}_{0}$}{
        \For{$j \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}
% TODO (later, let's not bother yet): replace x_i-c_i on Line 7 of Algorithm 2
% with a generalisation of x_i that can be an arbitrary summation/subtraction of
% domain sizes (propagating this change to the rest of the paper, particularly
% Assumption 1, the proof of Lemma 1)

@@

  Now, let us reason manually about the model count of $\phi$ when
  $\Delta = \emptyset$. Predicate $Q$ can only take one value, $Q = \emptyset$.
  The value of $P$ is fixed over $\Gamma' \times \Gamma'$ by \cref{eq:example2},
  but it is allowed to vary freely over
  $(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')$ since
  \cref{eq:example1} is vacuously satisfied by all structures. Therefore, the
  correct WFOMC should be $2^{|\Gamma|^2 - |\Gamma'|^2}$.

@@

We compare the algorithms on three benchmarks from previous work. The first is
the function-counting problem from \cref{example:functions}, previously examined
by \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
second is a variant of the well-known `Friends \& Smokers' Markov logic
network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In \Ctwo{},
\FO{}, and \UFO{}, it can be formulated as
\begin{gather*}
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land{}\\
  (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\end{gather*}
or, equivalently, in conjunctive normal form as
\begin{gather*}
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land{}\\
  (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\end{gather*}
Finally, we include the bijection-counting problem previously used by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Its
formulation in \FO{} is described in \cref{example:overall}. The equivalent
formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

@@

% TODO: run this section through editgpt and grammarly

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas.
Third, we have proposed an approach to compile function definitions into C++
programs with support for arbitrary-precision arithmetic. Lastly, we have
provided experimental evidence demonstrating that \Cranetwo{} can scale to much
larger domain sizes than \textsc{FastWFOMC} while handling more formulas than
\textsc{ForcLift}.

There are many potential avenues for future work. Specifically, a more thorough
experimental study is needed to understand how WFOMC algorithms compare in terms
of their ability to handle different formulas and their scalability with respect
to domain size. Additionally, further characterisation of the capabilities of
\Cranetwo{} can be explored. For example, \emph{completeness} could be proven
for a fragment of first-order logic such as \Ctwo{} (using a suitable encoding
of counting quantifiers). Moreover, the efficiency of a WFOMC algorithm in
handling a particular formula can be assessed using \emph{fine-grained
  complexity}. In the case of \textsc{Crane} and \Cranetwo{}, this can be done
by analysing the equations~\cite{DBLP:conf/kr/DilkasB23}. By doing so,
efficiency can be reasoned about in a more implementation-independent manner by
making claims about the maximum degree of the polynomial that characterises any
given solution.

>>>>

\section{Introduction}

@@

\emph{First-order model counting} (FOMC) is the task of counting the number of
models of a sentence in first-order logic over some given domain(s). The
(symmetric) weighted variant of this problem, known as WFOMC, seeks to compute
the total weight of the models~\cite{DBLP:conf/ijcai/BroeckTMDR11}. WFOMC
relates to its propositional predecessor weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08} and, more conceptually, to other
attempts to unify logic and probability such as probabilistic
logic~\cite{DBLP:journals/ai/Nilsson86}, fuzzy
logic~\cite{novak2012mathematical}, and graded logics~\cite{vsaletic2024graded}.
Moreover, WFOMC is a key approach to \emph{lifted (probabilistic) inference},
which aims to compute probabilities more efficiently by leveraging symmetries
inherent in the problem description~\cite{DBLP:conf/ecai/Kersting12}.

@@

Lifted inference is an active area of research, with recent work in domains such
as constraint satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23} and
probabilistic answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}.
WFOMC has been used for inference on probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. By considering domains of
increasing sizes, the FOMC of a formula can be seen as an integer sequence. FOMC
algorithms have been utilised for discovering new
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} as well as
conjecturing~\cite{DBLP:conf/ilp/BarvinekB0ZK21} and
constructing~\cite{DBLP:conf/kr/DilkasB23} recurrence relations and other
recursive structures that describe these sequences. Additionally, FOMC
algorithms have been extended to perform
\emph{sampling}~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}.

@@

The complexity of FOMC is typically characterised in terms of \emph{data
  complexity}. This involves fixing the formula and determining whether an
algorithm exists that can compute the FOMC in time polynomial with respect to
the domain size(s). If such an algorithm exists, the formula is called
\emph{liftable}~\cite{DBLP:conf/starai/JaegerB12}.
\citeauthor{DBLP:conf/pods/BeameBGS15}~\shortcite{DBLP:conf/pods/BeameBGS15}
demonstrated the existence of an unliftable formula with three variables. It is
also known that all formulas with up to two variables are
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}. The liftable
fragment of formulas with two variables has been expanded with various
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23}, counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21} and in other
ways~\cite{DBLP:conf/nips/KazemiKBP16}.

@@

There are various FOMC algorithms with different underlying principles. Perhaps
the most prominent class of FOMC algorithms is based on \emph{first-order
  knowledge compilation} (FOKC). In this approach, the formula is compiled into
a representation (such as a circuit or graph) by iteratively applying
\emph{compilation rules}. This representation can then be used to compute the
FOMC for any combination of domain sizes. Algorithms in this class include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its recent extension
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. The former compiles formulas into
circuits, while the latter compiles them first to graphs and then to (algebraic)
equations. Another FOMC algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, is based on cell enumeration.
Other algorithms utilise local search~\cite{DBLP:journals/pvldb/NiuRDS11},
junction trees~\cite{DBLP:conf/aaai/VenugopalSG15}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime approximation via
upper/lower bound construction~\cite{DBLP:conf/ijcai/BremenK20}.

@@

The recently proposed \textsc{Crane} algorithm marked significant progress in
handling formulas beyond the capabilities of \textsc{FastWFOMC} and
\textsc{ForcLift}, yet it fell short in critical aspects. \textsc{Crane} was
incomplete in that it could only construct function definitions, leaving users
to manually evaluate these functions to obtain model counts. This limitation
prevented \textsc{Crane} from serving as a convenient black box solution for
FOMC. Furthermore, it introduced recursive functions without defining necessary
base cases, adding further complexity to the user.

In this work, we present \Cranetwo{}, the breakthrough that not only addresses
these gaps but also pushes the boundaries of scalability to unprecedented
levels. With \Cranetwo{}, we eliminate the manual steps, providing a fully
automated FOMC algorithm. Moreover, \Cranetwo{} can scale to domain sizes over
\num{500000} times larger than the previous state of the art, redefining what is
possible in the field of first-order model counting.

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a FOMC
problem are compiled into C++ programs. Note that a solution to FOMC that uses
compilation to C++ has been considered before~\cite{DBLP:conf/kr/KazemiP16},
however, the extent of formulas that could be handled was limited. Finally,
\cref{sec:experiments} presents experiments results comparing \Cranetwo{} with
other state-of-the-art FOMC algorithms.

@@

In \cref{sec:logic}, we provide a summary of the basic principles of first-order
logic. Then, in \cref{sec:threelogics}, we formally define (W)FOMC and discuss
the distinctions between three variations of first-order logic that are utilised
for FOMC.\@ Finally, in \cref{sec:algebra}, we introduce the terminology used
to describe the output of the original \textsc{Crane} algorithm, i.e., functions
and equations that define them.

@@

\subsection{FOMC Algorithms and Their Logics}\label{sec:threelogics}

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in FOMC based on the following aspects:
    \begin{enumerate*}[label=(\roman*)]
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate*}
    Here:
    \begin{enumerate*}[label=(\roman*)]
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate*}
  }\label{tbl:logics}
\end{table*}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in FOMC:
\begin{enumerate*}[label=(\roman*)]
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by the most recent implementation
  of
  \textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}.
\end{enumerate*}
The notation we use to refer to each logic is standard in the case of \Ctwo{}
and \UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be
more specific in the case of \FO{}. All three logics are function-free, and
domains are always assumed to be finite. As usual, we presuppose the
\emph{unique name assumption}, which states that two constants are equal if and
only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

@@

While the focus of this work is on FOMC, we still define the weighted variant of
the problem as Skolemization relies on weights even for unweighted FOMC.

@@

It is worth noting that these base cases overlap when $l = m = 0$ but remain
consistent since $0^{0} = 1$. Generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the FOMC
of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
FOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{function call $f(\mathbf{y})$ on the right-hand side of an equation in $\mathcal{E}$}{\label{line:functioncall}
    $\mathbf{x} \gets \text{the parameters of $f$ in its definition}$\;
    \ForEach{$y_{i} \in \mathbf{y}$} {
      \uIf{$y_{i} \in \mathbb{N}_{0}$}{
        $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$\;
      }
      \ElseIf{$y_{i} = x_{i} - c_{i}$ for some $c_{i} \in \mathbb{N}_{0}$}{
        \For{$j \gets 0$ \KwTo $c_{i} - 1$}{\label{line:lim}
          $\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$\;\label{line:insert}
        }
      }
    }
  }
\end{algorithm}
% (later, let's not bother yet): replace x_i-c_i on Line 7 of Algorithm 2 with a
% generalisation of x_i that can be an arbitrary summation/subtraction of domain
% sizes (propagating this change to the rest of the paper, particularly
% Assumption 1, the proof of Lemma 1)

@@

  Now, let us reason manually about the model count of $\phi$ when
  $\Delta = \emptyset$. Predicate $Q$ can only take one value, $Q = \emptyset$.
  The value of $P$ is fixed over $\Gamma' \times \Gamma'$ by \cref{eq:example2},
  but it is allowed to vary freely over
  $(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')$ since
  \cref{eq:example1} is vacuously satisfied by all structures. Therefore, the
  correct FOMC should be $2^{|\Gamma|^2 - |\Gamma'|^2}$.

@@

We compare the algorithms on three benchmarks from previous work. The first is
the function-counting problem from \cref{example:functions}, previously examined
by \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
second is a variant of the well-known `Friends and Smokers' Markov logic
network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In \Ctwo{},
\FO{}, and \UFO{}, it can be formulated as
\begin{gather*}
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land{}\\
  (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\end{gather*}
or, equivalently, in conjunctive normal form as
\begin{gather*}
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land{}\\
  (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\end{gather*}
Finally, we include the bijection-counting problem previously used by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Its
formulation in \FO{} is described in \cref{example:overall}. The equivalent
formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

@@

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas.
Third, we have proposed an approach to compile function definitions into C++
programs with support for arbitrary-precision arithmetic. Lastly, we have
provided experimental evidence demonstrating that \Cranetwo{} can scale to much
larger domain sizes than \textsc{FastWFOMC} while handling more formulas than
\textsc{ForcLift}. Having FOMC algorithms that can efficiently handle large
domain sizes is especially crucial in the weighted setting. For example,
consider the `friends' instance examined in \cref{sec:experiments}, which models
a social network with friendships, smoking habits, and the probability of having
cancer. The utility of such a model would be significantly limited if
probabilities could only be efficiently computed for networks of at most
\num{1000} people.

There are many potential avenues for future work. Specifically, a more thorough
experimental study is needed to understand how FOMC algorithms compare in terms
of their ability to handle different formulas and their scalability with respect
to domain size. Additionally, further characterisation of the capabilities of
\Cranetwo{} can be explored. For example, \emph{completeness} could be proven
for a fragment of first-order logic such as \Ctwo{} (using a suitable encoding
of counting quantifiers). Moreover, the efficiency of a FOMC algorithm in
handling a particular formula can be assessed using \emph{fine-grained
  complexity}. In the case of \textsc{Crane} and \Cranetwo{}, this can be done
by analysing the equations~\cite{DBLP:conf/kr/DilkasB23}. By doing so,
efficiency can be reasoned about in a more implementation-independent manner by
making claims about the maximum degree of the polynomial that characterises any
given solution.
